Towards Proof-Reconstruction of Problems in Classical Propositional Logic

### 2017-10-18 paper/sections/proof-reconstruction.tex
[ paper ] minor changes.
[ paper ] lots of grammar fixes.

With the \TSTP  derivation from the Step~2, we process the derivation with the
\Athena translator tool~\cite{Athena}.
\Athena parses the \TSTP format, analyzes the
derivation and generates a representation of the natural deduction
proof using a tree data structure (see the properties
of this tree in Section~\ref{sssec:metis-proofs}).
We perform an analysis of the proof-tree to remove some
unnecessary steps that introduce redundancies or unused input for
\Athena.

As a result of using \Athena, we get an \Agda file, the proof-term. To
generate this file, \Athena has used names of functions and theorems from
\Agda libraries:
\texttt{agda-prop}~\cite{AgdaProp}, and \texttt{agda-metis}~\cite{AgdaMetis}.

Finally, we type-check the \Agda proof-term. If type-checking
succeed, the \TSTP derivation delivered by \Metis was correct.
In that case \Agda outputs an interface file.
Otherwise, when type-checking fails, the failure must be
investigated by the user, checking for the error in \Metis, \Athena,
or in \Agda.

In the remainder part of this section, a formal description using
type theory is presented to build definitions and theorems of functions
necessary to emulate \Metis' inference rules. As notation convention,
We write the type judgments with ``$a : A$'' to denote that $a$
has $\abbre{A}$ type. Types are written using small caps.
We use \Prop type for propositions and the dependent type, $Γ ⊢ φ$ for
theorems where $Γ$ has $\abbre{List}~\Prop$ type, and $φ$ has type $\Prop$.
$\abbre{List}$ type is the usual inductive type for lists. For syntactical
equivalence, we write $φ ≡ ψ$ for $φ : \Prop$ and $ψ : \Prop$.

@@

To emulate \Metis' inference rules in type theory,
we observed that transformations performed by some rules were given
by \emph{general recursive} functions.
But algorithms based on general recursions can not be translated
directly in type theory since it is not a guarantee they terminate.
For that reason, we follow the technique described
in~\cite{Bertot2004} to avoid termination problems by modifying the
recursive
functions to be \emph{structural recursive}.
% (see \cite{Coquand1992,Abel2002,Bove2005} for more details).

@@

General recursive calls can be translated into structurally recursive
calls by using for instance, the \emph{Bounded Recursion} technique
presented in~\cite{Bertot2004}. For a yet another method, we refer
the reader to \cite{Coquand1992,Abel2002,Bove2005}.
The former technique defines a new function based on the original
recursive function by adding an additional argument to the function.
The new argument is the \emph{bound}, a natural number given by the
function complexity. In other words, the added argument will store
the number of times the function needs to call itself to get the
expected outcome.

Henceforth, a natural number has \Nat type, and it is defined as
usual. That is, \zero and \suc are the data constructors and some
arithmetic operations are defined: ($+,\,-,\,*$). We include syntax
sugar for $\zero$, $\suc$, $\suc(\zero)$, $\cdots$, with the
classical representation: $1$, $2$, $\cdots$ as well.

Therefore, for a general recursive function $f: \Prop \to \Prop$
with termination problems, we states a new function
$f^{*} : \Prop \to \Nat \to \Prop$ where all recursive calls must be
done by using induction on the \Nat type structure.

@@

Recall, the \emph{complexity measure} defined in \cite{Agudelo-Agudelo2017}
and presented in \cite{VanDalen1994} as \emph{the rank for a
proposition}, is a function $\Prop \to \Nat$ that assigns a natural
number for a proposition $φ$ like the following $rank$ function.
The symbols, $c_{¬}$ and $c_{\square}$ is a non-negative integer
number called the complexity of the connective.

\begin{align*}
  \begin{split}
  % \label{eq:rank-definition}
    &rank : \Prop → \Nat\\
    &rank(φ) =
    \begin{cases}
    0, &\text{for atomic }φ \\
    rank(φ₁) + c_{¬},  &\text{ if } φ ≡ \neg φ₁ \\
    \max{(rank(φ₁),rank(φ₂))} + c_{\square},
      &\text{ if } φ ≡ φ₁\,\square\, φ₂.
    \end{cases}
  \end{split}
\end{align*}

We define the $unshunt_{cm}: \Prop → \NN$ function as
the complexity measure for $unshunt_0$ function.

\begin{align}
  \label{eq:unshunt-rank-definition}
  \begin{split}
      &unshunt_{cm} : \Prop → \Nat\\
      &unshunt_{cm}(φ) =\\
      &\hspace{3mm}
        \begin{cases}
        unshunt_{cm}(φ₃) + 2, &\text{ if }φ ≡ φ₁ ⇒ (φ₂ ⇒ φ₃)\\
        \max(unshunt_{cm}(φ₂),unshunt_{cm}(φ₃)) + 1,
          &\text{ if }  φ ≡ φ₁ ⇒ (φ₂ ∧ φ₃)\\
        0, &\text{ otherwise.}
        \end{cases}
  \end{split}
\end{align}

@@

For a more complete treatment of general recursion, and how to
define well-founded and well-structural recursive functions, we refer
the reader to Ref.~\cite{Bove2005}, where Bove and Capettra described
not only general recursion, but they also treat mutual recursive
function, among other kinds.

@@

In this section, we emulate each \Metis inference rule by proving a
function with the same name, and a associated theorem.
The following pattern is used to present the formalization.
The function name written in bold print
is a \Metis inference rule in \TSTP derivations.
The same function name written in italics refers to its emulated
version.
Each emulated rule states a theorem to prove its soundness.
This theorem is named by using the prefix \emph{thm}
and the name of the rule.

\begin{itemize}
\item \texttt{rule} is a $n$-ary \Metis inference rule
\item $rule$ is the emulated version function  for
the \texttt{rule} inference.
\begin{align*}
  \begin{split}
    &rule : \Prop → \Prop → \cdots → \Prop\\
    &rule(φ₁,\cdots, φₙ, ψ) =
      \begin{cases}
      ψ, &\text{ if }rule\text{ built }ψ\text{ by applying valid}\\
         &\text{ transformations or other rules to }φ\\
      φ, &\text{ otherwise.}
      \end{cases}
  \end{split}
\end{align*}

@@

To prove a goal, \Metis splits the goal into
disjoint cases. This process produces a list of new subgoals, the
conjunction of these subgoals implies the goal. Then, a proof of the
goal becomes in smaller proofs, one refutation for each subgoal.
These subgoals are introduced in the \TSTP derivation with the \strip
inference rule as we show with the following excerpt.

@@

To prove the correctness of the process mentioned above, first,
we need to establish some lemmas before introduce our main goal of this part,
Theorem~\ref{thm:thm-strip}.

@@

\begin{sketchproof} Use induction on the cases defined by the outcome of the
$unshunt$ function in Eq.~(\ref{eq:unshunt-structured}).
\begin{itemize}
	\item If $n = 0$, by definition we conclude $Γ ⊢ φ$.
	\item If $n = 1$, we apply the following theorem proved in~\cite{AgdaProp},
		\begin{equation*}
		\texttt{∧⇒-to-⇒⇒}\ :\  Γ ⊢ (φ₁ ∧ φ₂) ⇒ φ₃ → Γ ⊢ φ₁ ⇒ (φ₂ ⇒ φ₃).
		\end{equation*}
  \item For $n > 1$, we will use induction on the structure of the second
			  argument.
\begin{itemize}
\item $(φ₁ ⇒ (φ₂ ⇒ φ₃))$
\begin{equation*}
  \begin{bprooftree}
  \AxiomC{$Γ ⊢ unshunt(φ₁ ⇒ (φ₂ ⇒ φ₃), n)$}
  \RightLabel{axiom}
  \UnaryInfC{$Γ ⊢ unshunt(φ₁ ⇒ (φ₂ ⇒ φ₃), n-1)$}
  \RightLabel{lem-unshunt}
  \UnaryInfC{Γ ⊢ (φ₁ ∧ φ₂) ⇒ φ₃}
  \RightLabel{∧⇒-to-⇒⇒.}
  \UnaryInfC{Γ ⊢ φ₁ ⇒ (φ₂ ⇒ φ₃)}
  \end{bprooftree}
\end{equation*}

@@

Let us now define the $split$ function that yields the conjunction of subgoals that implies the goal of the problem in the \Metis \TSTP derivations.

@@

The definition above suffers termination problems since it is not
a structural recursion. Therefore, we have reformulated the equation
above to solve such an issue following the technique
described in Section~\ref{ssec:structural-recursion}. We refer to $split$
as the function after applying the process mentioned above here and below
unless otherwise stated.

@@

\item $(φ ≡ ¬ (φ₁ ∨ φ₂))$. We show that $Γ ⊢ ¬ φ₁$ and $Γ ⊢ ¬ φ₂$.
The result follows by introducing the conjunction and applying after
De Morgan Law.

@@

\begin{lemma}[subst]
  \label{lem:subst}
  Substitution theorem.
\begin{equation}
  \label{eq:substitution-theorem}
  \begin{bprooftree}
  \AxiomC{$Γ ⊢ φ$}   \AxiomC{$ψ ≡ φ$}
  \RightLabel{subst.}
  \BinaryInfC{$Γ ⊢ ψ$}
  \end{bprooftree}
\end{equation}
\end{lemma}

@@

\begin{theorem}[thm-strip]
\label{thm:thm-strip}
 If $s_i$ are the subgoals of the goal $φ$ for $i=1,\cdots,n$, and
 $\{ Γ ⊢ s_i \}_{i =1}^n$, then $Γ ⊢ φ$.
\end{theorem}

@@

Each proof of a subgoal is a refutation, thereby each proof assumes
the negation of its subgoal. The \negate rule
assumes the negation of such subgoals. This rule
appears after applying the \strip inference to the goal.

@@

Then, to emulate this rule, we just introduce the negation
of the subgoal by using $assume$ rule from the inferences in
Fig.~\ref{fig:CPL-inference-rules}.

@@

The \conjunct rule extracts from a
conjunction one of its conjuncts. This rule is a generalization of the
projection rules for the conjunction connective as the
following \TSTP excerpt shows us.

@@

\begin{definition}[conjunct]
\begin{align}
  \begin{split}
  &conjunct : \Prop \to \Prop \to \Prop\\
  &conjunct(φ, ψ) =
    \begin{cases}
        ψ, &\text{ if }φ ≡ ψ\\
        ψ, &\text{ if }φ ≡ φ₁ ∧ φ₂\text{ , }ψ ≡ conjunct(φ₁, ψ)\\
        ψ, &\text{ if }φ ≡ φ₁ ∧ φ₂\text{ , }ψ ≡ conjunct(φ₂, ψ)\\
        φ, &\text{ otherwise.}
      \end{cases}
  \end{split}
\end{align}
\end{definition}

\begin{theorem}[thm-conjunct]
  \label{thm:thm-conjunct}
  $Γ ⊢ φ$, and $ψ  : \Prop$ , then $Γ ⊢ conjunct(φ, ψ)$.
\end{theorem}
\begin{proof}\hspace{2cm}
\begin{itemize}
  \item For the case, $φ ≡ ψ$, $Γ ⊢ conjunct(φ, ψ)$ normalizes to $Γ ⊢ ψ$.
Then, we get the desire conclusion by applying the $subst$ theorem.
  \item If the proposition $φ$ is a conjunction, and we can get $ψ ≡ conjunct(φ_{i}, ψ)$ for some $i = 1,\ 2$, then,

\begin{equation}
  \begin{bprooftree}
  \AxiomC{$Γ ⊢ φ₁ ∧ φ₂$}
  \RightLabel{∧-proj$_{i}$}
  \UnaryInfC{$Γ ⊢ φ_{i}$}
  \UnaryInfC{$Γ ⊢ conjunct(φ_{i}, ψ)$}
  \AxiomC{$ψ ≡ conjunct(φ_{i}, ψ)$}
  \RightLabel{subst.}
  \BinaryInfC{$Γ ⊢ ψ$}
  \end{bprooftree}
\end{equation}
\item The last case follows trivially by using the same hypothesis.
\end{itemize}
\end{proof}

@@

Logic equivalence between propositions is a major issue to justify
prover reasoning steps. Since we left out semantics to treat only the
syntax aspects of the propositions, our approach shows equivalence
by converting propositions to their conjunctive normal forms.
But also, reordering the conjunctions and the inner disjunctions into the
formulas to match them.
In this section, we provide some theorems to perform such reordering tasks,
the goals are the following.

\begin{itemize}
  \item Reorder a disjunction $φ = φ₁ ∨ φ₂ ∨ \cdots φₙ$ to match with
$ψ = ψ₁ ∨ ψ₂ ∨ \cdots ∨ ψₙ$, when for all $i = 1, \cdots, n$, exists at
least one $ψⱼ$ such that $φᵢ ≡ ψⱼ$ for $j = 1, \cdots, n$.
  \item Reorder a CNF formula, $φ = φ₁ ∧ φ₂ ∧ \cdots$, to match with another
CNF formula $ψ = ψ₁ ∧ ψ₂ ∧ \cdots$ when for all $i = 1, \cdots, n$, exists at
least one $ψⱼ$ such that $φᵢ ≡ ψⱼ$ for $j = 1, \cdots, n$.
\end{itemize}

First, we define the $rdisj$ function to convert
a disjunction in its right-associative form.
The $rconj$ function converts conjunction to its right-associative form is not
presented here, but is defined in a similar way (see its implementation
in~\cite{AgdaMetis}).

@@

The $build_{∨}$ function in Eq.~(\ref{eq:build-or}) intends to construct a
disjunction from another disjunction, specifically, this functions will try
to rearrange the disjuncts in the source formula to match with the target
disjunction formula, similar as we described above.

@@

From now, we assume all propositions are right-associative unless otherwise stated.
The $factor$ function in Eq.~(\ref{eq:factor-definition}) intends to simplify a
special case of disjunction, the repeated disjuncts (e.g., $factor(φ ∨ φ) \mapsto\ φ$).
Notice that other cases like $φ ∨ (ψ ∨ φ)$ do no reduce to $(ψ ∨ φ)$.

@@

To construct a disjunction $ψ$ from another disjunction $φ$, we have used ideas
from the description in \citeauthor{bohme2010}~\cite{bohme2010} to prove equality
between nested disjunctions. The result is the $sbuild_{∨}$ function that uses
every disjunct from the source formula, $φ$, to build up the target disjunction $ψ$.

\begin{definition}[sbuild$_{∨}$]
\begin{align}
  \begin{split}
  \label{eq:strong-build-or-definition}
    &sbuild_{∨} : \Prop → \Prop → \Prop\\
    &sbuild_{∨}(φ, ψ) =\\
    &\hspace{3mm}
    \begin{cases}
      factor(build_{∨}(φ₁, ψ) ∨ build_{∨}(φ₂,ψ)),  &\text{ if }φ≡ φ₁ ∨ φ₂\\
      build_{∨}(φ, ψ),  & \text{ otherwise.}
    \end{cases}
  \end{split}
\end{align}
\end{definition}

@@

\begin{example}
Let see the main idea behind the lem-sbuild$_{∨}$ lemma.
We will build the disjunction $((p ∨ q) ∨ r)$ from the disjunction
$(r ∨ (q ∨ p))$.
\begin{equation*}
  \begin{bprooftree}
  \AxiomC{$Γ ⊢ q$}
  \RightLabel{∨-intro₂}
  \UnaryInfC{$Γ ⊢ p ∨ q$}
  \RightLabel{∨-intro₁}
  \UnaryInfC{$Γ ⊢ (p ∨ q) ∨ r$}

@@

\begin{definition}[reorder$_{∧}$]
  \begin{align*}
      \begin{split}
        &reorder : \Prop → \Prop \to \Prop\\
        &reorder_{∧}(φ, ψ) =
        \begin{cases}
          φ, &\text{ if }φ ≡ ψ\\
          ψ₁ ∧ ψ₂, &\text{ if } ψ ≡ ψ₁ ∧ ψ₂\text{, }reorder_{∧}(φ, ψ₁) ≡ ψ₁\\
                  &\text{and }reorder_{∧}(φ, ψ₂)≡ ψ₂\\
          φ,       &\text{ if } ψ ≡ ψ₁ ∧ ψ₂\\
          conjunct(φ, ψ), &\text{ otherwise.}
        \end{cases}
      \end{split}
  \end{align*}
\end{definition}

\begin{lemma}[lem-reorder$_{∧}$]
  \label{lem:lem-reorder-and}
  $Γ ⊢ φ$, and $ψ : \Prop$ then $Γ ⊢ reorder(φ, ψ)$.
\end{lemma}

@@

To emulate this rule, we perform rearrangements inside the formulas
involved to match with the pattern expected by the resolution theorem
described above. We begin by defining a convenient version of
the binary resolution in Def.~\ref{def:rsol}.

@@

In Def.~\ref{def:resolve}, the emulated rule for $resolve$
has four arguments. The first three were described
above. The last one, ψ, represents the expected conclusion in the \Metis
derivation.

@@

In this section we provide the functions and theorems to
remove inside a normalized formula the definitions showed in
Fig.~\ref{fig:conjunctive-disjunctive-simpl}.

@@

Now, we have removed redundancies in the disjunctions by applying
$rm_{∨}$ function, we can define a similar function to work with
conjunctions, the $rm_{∧}$ function.

@@

Continuing with the rest of tautologies from
Fig.~\ref{fig:conjunctive-disjunctive-simpl}, we define
$ndisj_{∨}$ function and $ndisj_{∨}$ function that
take into account the following cases.

@@

\begin{remark}
A similar treatment defines the $canon_{⊻} : \NProp \to \NProp$
function that removes redundancies related with the exclusive
disjunction connective.
% Its theorem, thm-canon$_{⊻}$ is presented in~\cite{AgdaMetis}.
We left out this description for the brevity of this
document and refer the interested reader to \cite{AgdaMetis} for
more details.
\end{remark}

Finally, we can translate formulas from \Prop to \NProp by defining
the $nnf₀$ function. We have taken
inspiration on the $nnf$ function described in
\citeauthor{Bezem2002}~\cite{Bezem2002}. Their $nnf$ function
translates formulas to their classical negative normal form. In the
reference, the authors avoid a termination problem with a similar
function by using as an extra argument, the polarity of the formula.
We follow the same idea in Def.~\ref{def:nnf-zero}.
Therefore, we need the \abbre{Polarity} type with
two constructors: $⊕$ to denote positive polarity, and $⊖$ to denote
negative polarity of the formula.

@@

Since all transformations performed above come from equivalences
in classical propositional logic, we have stated the following theorem with a purpose for the next section.

>>>>

With the \TSTP  derivation from Step~2, we process the derivation with the
\Athena translator tool~\cite{Athena}.
\Athena parses the \TSTP format, analyzes the
derivation and generates a representation of the natural deduction
proof using a tree data structure (see the properties
of this tree in Section~\ref{sssec:metis-proofs}).
We perform an analysis of the proof-tree to remove some
unnecessary steps that introduce redundancies or unused input for
\Athena.

As a result of using \Athena, we get an \Agda file, the proof-term. To
generate this file, \Athena uses names of functions and theorems from
\Agda libraries:
\texttt{agda-prop}~\cite{AgdaProp}, and \texttt{agda-metis}~\cite{AgdaMetis}.

Finally, we type-check the \Agda proof-term. If the type-checking
success, the \TSTP derivation delivered by \Metis is correct.
In that case \Agda outputs an interface file.
Otherwise, when type-checking fails, the failure must be
investigated by the user, checking for the error in \Metis, \Athena,
or in \Agda.

In the remainder part of this section, a formal description using
type theory is presented to build definitions and theorems of functions
necessary to emulate \Metis' inference rules. As a convention,
We write the type judgments as ``$a : A$'' to denote that $a$
has $\abbre{A}$ type. Types are written using small caps.
We use \Prop type for propositions and the dependent type, $Γ ⊢ φ$ for
theorems where $Γ$ has $\abbre{List}~\Prop$ type, and $φ$ has type $\Prop$.
$\abbre{List}$ type is the usual inductive type for lists. For syntactical
equivalence, we write $φ ≡ ψ$ for $φ : \Prop$ and $ψ : \Prop$.

@@

To emulate \Metis' inference rules in type theory,
we observed that transformations performed by some rules are given
by \emph{general recursive} functions.
Algorithms based on general recursions can not be translated
directly in type theory since it is not a guarantee they terminate.
For that reason, we follow the technique described
in~\cite{Bertot2004} to avoid termination problems by modifying the
recursive
functions to be \emph{structural recursive}.
% (see \cite{Coquand1992,Abel2002,Bove2005} for more details).

@@

General recursive calls can be translated into structurally recursive
calls by using for instance, the \emph{Bounded Recursion} technique
presented in~\cite{Bertot2004}. For a yet another method, we refer
the reader to \cite{Coquand1992,Abel2002,Bove2005}.
The former technique defines a new function based on the original
recursive function by adding an argument to the function.
The new argument is the \emph{bound}, a natural number given by the
function complexity. In other words, the added argument will store
the number of times the function needs to call itself to get the
expected outcome.

Henceforth, a natural number has \Nat type, and it is defined as
usual, i.e., \zero and \suc are the data constructors and some
arithmetic operations are defined: ($+,\ -,\ *$). We include syntax
sugar for $\zero$, $\suc$, $\suc(\zero)$, $\cdots$, with the
classical representation: $1$, $2$, $\cdots$ as well.

Therefore, for a general recursive function $f: \Prop \to \Prop$
with termination problems, we state a new function
$f^{*} : \Prop \to \Nat \to \Prop$ where all recursive calls must be
done by using induction on the \Nat type structure.

@@

Recall the \emph{complexity measure} defined in \cite{Agudelo-Agudelo2017}
and presented in \cite{VanDalen1994} as \emph{the rank for a
proposition}, is a function $\Prop \to \Nat$ that assigns a natural
number for a proposition $φ$ like the following $rank$ function.
The symbols $c_{¬}$ and $c_{\square}$ are non-negative integer
numbers called the complexity of the respective connective.

\begin{align*}
  \begin{split}
  % \label{eq:rank-definition}
    &rank : \Prop → \Nat\\
    &rank(φ) =
    \begin{cases}
    0, &\text{for atomic }φ \\
    rank(φ₁) + c_{¬},  &\text{ if } φ ≡ \neg φ₁ \\
    \max{(rank(φ₁),\ rank(φ₂))} + c_{\square},
      &\text{ if } φ ≡ φ₁\,\square\, φ₂.
    \end{cases}
  \end{split}
\end{align*}

We define the $unshunt_{cm}: \Prop → \NN$ function as
the complexity measure for the $unshunt_0$ function.

\begin{align*}
  \label{eq:unshunt-rank-definition}
  \begin{split}
      &unshunt_{cm} : \Prop → \Nat\\
      &unshunt_{cm}(φ) =\\
      &\hspace{3mm}
        \begin{cases}
        unshunt_{cm}(φ₃) + 2, &\text{ if }φ ≡ φ₁ ⇒ (φ₂ ⇒ φ₃)\\
        \max(unshunt_{cm}(φ₂),\ unshunt_{cm}(φ₃)) + 1,
          &\text{ if }  φ ≡ φ₁ ⇒ (φ₂ ∧ φ₃)\\
        0, &\text{ otherwise.}
        \end{cases}
  \end{split}
\end{align*}

@@

For a more complete treatment of general recursion, and how to
define well-founded and well-structural recursive functions, we refer
the reader to Ref.~\cite{Bove2005}, where Bove and Capettra described
not only general recursion, but they also treat mutual recursive
function, among others.

@@

In this section, we emulate each \Metis inference rule by proving a
function with the same name, and a associated theorem.
The following pattern is used to present the formalization.
The function name written in bold print
is a \Metis inference rule in \TSTP derivations.
The same function name written in italics refers to its emulated
version.
Each emulated rule states a respective theorem that proves its soundness.
This theorem is named by using the prefix \emph{thm}
and the name of the rule.

\begin{itemize}
\item \texttt{rule} is a $n$-ary \Metis inference rule
\item $rule$ is the emulated version function  for
the \texttt{rule} inference.
\begin{align*}
  \begin{split}
    &rule : \Prop → \Prop → \cdots → \Prop\\
    &rule(φ₁,\ \cdots, φₙ, ψ) =
      \begin{cases}
      ψ, &\text{ if }rule\text{ built }ψ\text{ by applying valid}\\
         &\text{ transformations or other rules to }φ\\
      φ, &\text{ otherwise.}
      \end{cases}
  \end{split}
\end{align*}

@@

To prove a goal, \Metis splits the goal into
disjoint cases. This process produces a list of new subgoals, the
conjunction of these subgoals implies the goal. Then, a proof of the
goal becomes, in smaller proofs, one refutation for each subgoal.
These subgoals are introduced in the \TSTP derivation with the \strip
inference rule as we show with the following excerpt.

@@

To prove the correctness of the process mentioned above,
we need to establish some lemmas before introduce our main goal of this part,
Theorem~\ref{thm:thm-strip}.

@@

\begin{sketchproof} Use induction on the cases defined by the outcome of the
$unshunt$ function in Eq.~(\ref{eq:unshunt-structured}).
\begin{itemize}
	\item If $n = 0$, by definition we conclude $Γ ⊢ φ$.
	\item If $n = 1$, we apply the following theorem proved in~\cite{AgdaProp},
		\begin{equation*}
		\texttt{∧⇒-to-⇒⇒}\ :\  Γ ⊢ (φ₁ ∧ φ₂) ⇒ φ₃ → Γ ⊢ φ₁ ⇒ (φ₂ ⇒ φ₃).
		\end{equation*}
  \item For $n > 1$, we use induction on the structure of the second
			  argument.
\begin{itemize}
\item $(φ₁ ⇒ (φ₂ ⇒ φ₃))$
\begin{equation*}
  \begin{bprooftree}
  \AxiomC{$Γ ⊢ unshunt(φ₁ ⇒ (φ₂ ⇒ φ₃), n)$}
  \RightLabel{axiom}
  \UnaryInfC{$Γ ⊢ unshunt(φ₁ ⇒ (φ₂ ⇒ φ₃), n-1)$}
  \RightLabel{lem-unshunt}
  \UnaryInfC{Γ ⊢ (φ₁ ∧ φ₂) ⇒ φ₃}
  \RightLabel{∧⇒-to-⇒⇒.}
  \UnaryInfC{Γ ⊢ φ₁ ⇒ (φ₂ ⇒ φ₃)}
  \end{bprooftree}
\end{equation*}

@@

We now define the $split$ function that yields the conjunction of subgoals
that implies the goal of the problem in the \Metis \TSTP derivations.

@@

The definition above has termination problems since it is not
a structural recursion. Therefore, we have reformulated the equation
above to solve such an issue following the technique
described in Section~\ref{ssec:structural-recursion}. We refer to $split$
as the function after applying the process mentioned above here and below
unless otherwise stated.

@@

\item $(φ ≡ ¬ (φ₁ ∨ φ₂))$. We show that $Γ ⊢ ¬ φ₁$ and $Γ ⊢ ¬ φ₂$.
The result follows by introducing the conjunction and applying
De Morgan Laws.

@@

\begin{lemma}[subst]
  \label{lem:subst}
  Substitution theorem.
\begin{equation*}
  \label{eq:substitution-theorem}
  \begin{bprooftree}
  \AxiomC{$Γ ⊢ φ$}   \AxiomC{$ψ ≡ φ$}
  \RightLabel{subst.}
  \BinaryInfC{$Γ ⊢ ψ$}
  \end{bprooftree}
\end{equation*}
\end{lemma}

@@

\begin{theorem}[thm-strip]
\label{thm:thm-strip}
 If $s_i$ are the subgoals of the goal $φ$ for $i=1,\cdots,n$ and
 $\{ Γ ⊢ s_i \}_{i =1}^n$, then $Γ ⊢ φ$.
\end{theorem}

@@

Each proof of a subgoal is a refutation, thereby each proof assumes
the negation of its subgoal. The \negate rule
assumes the negation of such subgoals, and
appears after applying the \strip inference to the goal.

@@

To emulate this rule, we introduce the negation
of the subgoal by using $assume$ rule from the inferences in
Fig.~\ref{fig:CPL-inference-rules}.

@@

The \conjunct rule extracts from a
conjunction one of its conjuncts. This rule is a generalization of the
projection rules for the conjunction connective as the
following \TSTP excerpt shows.

@@

\begin{definition}[conjunct]
\begin{align*}
  \begin{split}
  &conjunct : \Prop \to \Prop \to \Prop\\
  &conjunct(φ, ψ) =
    \begin{cases}
        ψ, &\text{ if }φ ≡ ψ\\
        ψ, &\text{ if }φ ≡ φ₁ ∧ φ₂\text{ , }ψ ≡ conjunct(φ₁, ψ)\\
        ψ, &\text{ if }φ ≡ φ₁ ∧ φ₂\text{ , }ψ ≡ conjunct(φ₂, ψ)\\
        φ, &\text{ otherwise.}
      \end{cases}
  \end{split}
\end{align*}
\end{definition}

\begin{theorem}[thm-conjunct]
  \label{thm:thm-conjunct}
  $Γ ⊢ φ$, and $ψ  : \Prop$ , then $Γ ⊢ conjunct(φ, ψ)$.
\end{theorem}
\begin{proof}\hspace{2cm}
\begin{itemize}
  \item For the case $φ ≡ ψ$, $Γ ⊢ conjunct(φ, ψ)$ normalizes to $Γ ⊢ ψ$.
Then, we get the desire conclusion by applying the $subst$ theorem.
  \item If the proposition $φ$ is a conjunction, and we can get $ψ ≡ conjunct(φ_{i}, ψ)$ for some $i = 1,\ 2$, then,

\begin{equation*}
  \begin{bprooftree}
  \AxiomC{$Γ ⊢ φ₁ ∧ φ₂$}
  \RightLabel{∧-proj$_{i}$}
  \UnaryInfC{$Γ ⊢ φ_{i}$}
  \UnaryInfC{$Γ ⊢ conjunct(φ_{i}, ψ)$}
  \AxiomC{$ψ ≡ conjunct(φ_{i}, ψ)$}
  \RightLabel{subst.}
  \BinaryInfC{$Γ ⊢ ψ$}
  \end{bprooftree}
\end{equation*}
\item The last case follows trivially by using the same hypothesis.
\end{itemize}
\end{proof}

@@

Logic equivalence between propositions is a major issue to justify
prover reasoning steps. Since we left out semantics to treat only the
syntax aspects of the propositions, our approach shows equivalence
by converting propositions to their conjunctive normal forms,
and reordering those and the inner disjunctions  to match them
In this section, we provide some theorems to perform such reordering tasks,
the goals are the following.

\begin{itemize}
  \item Reorder a disjunction $φ = φ₁ ∨ φ₂ ∨ \cdots ∨ φₙ$ to match with
$ψ = ψ₁ ∨ ψ₂ ∨ \cdots ∨ ψₙ$. For all $i$, there exists at least one $j$ such that $φᵢ ≡ ψⱼ$.
  \item Reorder a CNF formula, $φ = φ₁ ∧ φ₂ ∧ \cdots$, to match with another
CNF formula $ψ = ψ₁ ∧ ψ₂ ∧ \cdots$. For all $i$, there exists at least one $j$ such that $φᵢ ≡ ψⱼ$.
\end{itemize}

First, we define the $rdisj$ function to convert
a disjunction in its right-associative form.
The $rconj$ function that converts conjunction to its right-associative form
is not presented here, but is defined in a similar way (see its implementation
in~\cite{AgdaMetis}).

@@

The $build_{∨}$ function in Eq.~(\ref{eq:build-or}) intends to construct a
disjunction from another disjunction, specifically, this functions will try
to rearrange the disjuncts in the source formula to match with the target
disjunction formula, similar as we describe above.

@@

From now, we assume all propositions to be right-associative unless otherwise stated.
The $factor$ function in Eq.~(\ref{eq:factor-definition}) intends to simplify a
special case of disjunction, the repeated disjuncts (e.g., $factor(φ ∨ φ) \mapsto\ φ$).
Notice that other cases like $φ ∨ (ψ ∨ φ)$ do no reduce to $(ψ ∨ φ)$.

@@

To construct a disjunction $ψ$ from a disjunction $φ$, we have used ideas
from the description in \citeauthor{bohme2010}~\cite{bohme2010} to prove equality
between nested disjunctions. The result is the $sbuild_{∨}$ function that uses
every disjunct from the source formula, $φ$, to build up the target disjunction $ψ$.

\begin{definition}[sbuild$_{∨}$]
\begin{align*}
  \begin{split}
  \label{eq:strong-build-or-definition}
    &sbuild_{∨} : \Prop → \Prop → \Prop\\
    &sbuild_{∨}(φ, ψ) =\\
    &\hspace{3mm}
    \begin{cases}
      factor(build_{∨}(φ₁, ψ) ∨ build_{∨}(φ₂,ψ)),  &\text{ if }φ≡ φ₁ ∨ φ₂\\
      build_{∨}(φ, ψ),  & \text{ otherwise.}
    \end{cases}
  \end{split}
\end{align*}
\end{definition}

@@

\begin{example}
Let see the main idea behind the lem-sbuild$_{∨}$ lemma.
We build the disjunction $((p ∨ q) ∨ r)$ from the disjunction
$(r ∨ (q ∨ p))$.
\begin{equation*}
  \begin{bprooftree}
  \AxiomC{$Γ ⊢ q$}
  \RightLabel{∨-intro₂}
  \UnaryInfC{$Γ ⊢ p ∨ q$}
  \RightLabel{∨-intro₁}
  \UnaryInfC{$Γ ⊢ (p ∨ q) ∨ r$}

@@

\begin{definition}[reorder$_{∧}$]
  \begin{align*}
      \begin{split}
        &reorder_{∧} : \Prop → \Prop \to \Prop\\
        &reorder_{∧}(φ, ψ) =
        \begin{cases}
          φ, &\text{ if }φ ≡ ψ\\
          ψ₁ ∧ ψ₂, &\text{ if } ψ ≡ ψ₁ ∧ ψ₂\text{, }reorder_{∧}(φ, ψ₁) ≡ ψ₁\\
                  &\text{and }reorder_{∧}(φ, ψ₂)≡ ψ₂\\
          φ,       &\text{ if } ψ ≡ ψ₁ ∧ ψ₂\\
          conjunct(φ, ψ), &\text{ otherwise.}
        \end{cases}
      \end{split}
  \end{align*}
\end{definition}

\begin{lemma}[lem-reorder$_{∧}$]
  \label{lem:lem-reorder-and}
  $Γ ⊢ φ$, and $ψ : \Prop$ then $Γ ⊢ reorder_{∧}(φ, ψ)$.
\end{lemma}

@@

To emulate this rule, we perform a sequence of rearrangements inside the
formulas involved to match with the pattern expected by the resolution theorem
described above. We begin by defining a convenient version of
the binary resolution in Def.~\ref{def:rsol}.

@@

In Def.~\ref{def:resolve}, the emulated rule for $resolve$
has four arguments. The first three were described
above, the last one, ψ, represents the expected conclusion in the \Metis
derivation.

@@

In this section we provide functions and theorems to remove, inside a normalized formula, the definitions showed in
Fig.~\ref{fig:conjunctive-disjunctive-simpl}.

@@

Now, we have removed redundancies in the disjunctions by applying
$rm_{∨}$ function, we can define a similar function to work with
conjunctions, i.e., $rm_{∧}$.

@@

Continuing with the rest of tautologies from
Fig.~\ref{fig:conjunctive-disjunctive-simpl}, we define the
$ndisj_{∨}$ function and $ndisj_{∨}$ function that
take into account the following cases.

@@

\begin{remark}
A similar treatment defines the $canon_{⊻} : \NProp \to \NProp$.
This function removes redundancies related with the exclusive
disjunction connective.
% Its theorem, thm-canon$_{⊻}$ is presented in~\cite{AgdaMetis}.
We left out this description for the brevity of this
document and refer the interested reader to \cite{AgdaMetis} for
more details.
\end{remark}

Finally, we can translate formulas from \Prop to \NProp by defining
the $nnf₀$ function. This function is based on the $nnf$ function described in
\citeauthor{Bezem2002}~\cite{Bezem2002}. Their $nnf$ function
translates formulas to their classical negative normal form. In the
reference, the authors avoid a termination problem with a similar
function by using, as an extra argument, the polarity of the formula.
We follow the same idea in Def.~\ref{def:nnf-zero}.
Therefore, we need the \abbre{Polarity} type with
two constructors: $⊕$ to denote positive polarity, and $⊖$ to denote
negative polarity of the formula.

@@

Since all the previous transformations came from  equivalences in  the
classical propositional logic, the following theorem plays an important role
in the next section.
